---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

The dataset used for this project is from the extact of the General Social Survey ([GSS](http://gss.norc.org)) Cumulative File 1972-2012. According to the codebook, a random sampling strategy was used therefore we can generalize the results to the US population

In addition, The study can be categorized as a random sampling observational study. Therefore, it will be difficult to draw causal conclusions based on the datasets since there is no random assignments or any experiments involved.

## Part 2: Research question

**Are older generations generally more opposed to homosexual sex relations than younger generations based on the dataset? **

I am interested to test the general public conception that older generations are more conservative and more against homosexual sex relations

* * *

## Part 3: Exploratory data analysis


* * *

We will use two columns to build my hypothesis for the research question  

+ **age**
+ **homosexual**

The levels of response for the question "What about sexual relations between two adults of the same sex?" is as follows

```{r}
levels(gss$homosex)
```
We will group people who answered either "Always Wrong" or "Almst Always Wrg" as people who are strongly against homosexual relations, and use the proportion **__p__** for the hypothesis test

```{r}
sub_df <- gss[, c("age", "homosex")]
sub_df <- sub_df[complete.cases(sub_df), ] # remove NA values

sub_df <- mutate(sub_df, against_homo=factor(ifelse(homosex=="Always Wrong" |homosex=="Almst Always Wrg", "yes", "No")))
summary(sub_df)
```
I will group the people into two age group based on the median age value
```{r}
sub_df <- mutate(sub_df, age_group=factor(ifelse(age>median(age), "older", "younger")))
summary(sub_df)
```
Group the data by people's age group

```{r}
older <- sub_df[sub_df$age_group=="older", ]
younger <- sub_df[sub_df$age_group=="younger", ]
```
Construct a 95% confidence interval for the proportion of against-homo in the **older** generation

```{r}
inference(y = against_homo, data = older, statistic = "proportion", type = "ci", method = "theoretical", success = "yes")
```  

Next construct the 95% confidence interval for the proportion of against-homo in the **younger** generation
```{r}
inference(y = against_homo, data = younger, statistic = "proportion", type = "ci", method = "theoretical", success = "yes")
```


## Part 4: Inference

For this research question, we would like to make conclusions about the difference in the two population proportions. We will construct our hypothesis as follows:
$$H_0: p_{old} - p_{young} = 0$$
$$H_0: p_{old} - p_{young} >0$$
where p is the proportion of people who are against homosexual sex relations. 
Let's start by check the conditions for using the normal model to analyse the result. 

- The sample observations are independent
    * A random sampling was used and sample size is smaller than 10% of the population 
- success-failure condition
    * First calculate the pooled proportion of people who are against homosexual
    ```{r}
    stats <- sub_df %>%
      group_by(against_homo) %>%
        summarise(ct=n()) %>%
          mutate(percentage=ct/sum(ct))
    pool <- stats[stats$against_homo=="yes", ]$percentage
    ```
    * Then check if expected success/failure count is larger than 10
    ```{r}
    total <- count(sub_df)
    success <- pool*total
    failure <- (1-pool)*total
    success
    failure
    ```  
    
All conditions are met for hypothesis test based on the above result. 

**First**, we can calculate the 95% condidence intervals for difference of two proportions between the two age groups
```{r}
inference(y = against_homo, x=age_group, data = sub_df, statistic = "proportion", type = "ci", method = "theoretical", success = "yes", alternative = "greater", boot_method = "se")
```  

We are 95% confident that the differnece of disapproval rate between older and younger generations is between 11.5% to 13.5%.


**Second**, we will continue out hypothesis test to see if we reject or fail to reject the null hypothesis

```{r}
inference(y = against_homo, x=age_group, data = sub_df, statistic = "proportion", type = "ht", method = "theoretical", success = "yes", alternative = "greater")
```
From the above result, we have a **p-value** less than 0.0001. Therefore, we will reject the null hypothesis and conclude that older generations are more opposed to homosexual sex relation than younger generations.